Constructing Sailing Match Race Schedules: Round-Robin Pairing Lists

### 2015-06-19 paper.tex
Removed vspace and modified subsection on orientation
no more negative space
in 3.4 orient[2,0] = 1 and orient[3,0] = 0 due to criterion 8

\section{Introduction}
\vspace{-2mm}
This work describes a round-robin competition format that arises from sailing, known as match
racing. Competitors, i.e.\ \emph{skippers}, compete against each other in a series of matches taking
place in rounds, known as \emph{flights}.  A match is composed of two skippers, with each skipper in
a boat on their own. Skippers in a match set off together, one on the port side, the other starboard,
and first home wins that match. Criteria published by the International Sailing Federation (ISAF)
dictate what makes a legal match race schedule and what should be optimized. This is a rich set of
criteria, 13 in all, and as far as we know, all published schedules have been produced by hand. This
is a daunting task. A close inspection of these schedules reveals that most are illegal, violating
many of the match race scheduling criteria, many schedules are missing, and some that are legal are
far from the ISAF definition of optimality.

@@

\section{Problem Definition: Round-Robin Pairing Lists}\label{sec:definition}
\vspace{-2mm}
The guidelines for running match racing events~\cite{isaf} places a number of criteria on how the
competing skippers are scheduled into flights and matches, known as pairing lists. Several of these
criteria are applicable when the number of skippers exceeds the number of available boats, in which
case skippers have to change in and out of boats, and are allowed time in the schedule to check and
fine-tune the boat they change into. Typically, matches set off at 5 minute intervals with each
match taking about 20 minutes. Therefore, if we have 10 skippers and 10 boats we have 45 matches,
9 flights each of 5 matches, and if each flight takes about 50 minutes,  eight or nine
flights a day is a reasonable achievement for most events~\cite{isaf}. Consequently, the number of
boats and skippers involved is typically small.

@@

\section{The Constraint Models}\label{sec:models}
\vspace{-2mm}
Our approach produces schedules in four stages. The first stage produces a schedule that respects
criteria~1, 4, 11 and~12, and minimizes criteria~10 (boat changes). The second stage constructs a
schedule that respects criteria~1, 4, 11, 12 (again), has criterion~10 as a hard constraint generated
from first stage, and minimizes criterion~6 (balance). This results in a multi-objective schedule that attempts to
minimize boat changes \emph{and} balance. The third stage satisfies criterion~9, and is a simple
translation of the schedule produced in stage 2.  The final stage orients skippers within matches to
address criteria~2, 3, 5 and~8.

@@

\subsection{Stage 1: Minimizing Boat Changes}
\vspace{-1mm}
\paragraph{Modeling skippers:} The first thing we model, in the first half of
Figure~\ref{model:stage1}, is a skipper. Each skipper $\sigma$ has both a temporal view of their
schedule (``who, if anyone, am I racing in the match at time $t$?''), and a state view (``where am
I racing within flight $f$?'').  The temporal view is via the array $\timeSlot$ defined in
\eqref{var:timeSlot}. If variable $\timeSlot[{i}] = k$ and $k \geq 0$ then this skipper is in a
match with skipper~$k$ at time~$i$.

@@

\subsection{Stage 2: Minimizing Imbalances}
\vspace{-1mm}
Having produced a schedule that minimizes boat changes, we then minimize imbalance (i.e.\ apply
criterion~6) by extending the model in Figure~\ref{model:stage2}.  Assuming the first stage
produces a schedule with $\beta$ boat changes we post this as a hard constraint \eqref{C11} in stage
2. For each skipper $\sigma$ we introduce a zero/one variable $position[{i,j}]$ \eqref{V9} which is
1 if and only if skipper $\sigma$ is in a match in time $m \cdot j + i$, i.e.\ in the $i^{th}$ match
of the  $j^{th}$ flight \eqref{C12}. Imbalance \eqref{V10} is then computed for each position, as
the absolute amount over or under the expected presence in a given position \eqref{C13}. Finally, we
compute the maximum imbalance over all positions for this skipper (\ref{V11}, \ref{C14}).

@@

\subsection{Stage 3: Renaming Skippers}
\vspace{-1mm}
Stage 3 then takes as input the schedule produced in stage 2 and renames skippers in the last round
of the last flight to be $(0,1)$, satisfying criterion~9. This is a simple linear process. The
schedule produced for 7 skippers and 6 boats is shown in Table~\ref{tab2}.

\subsection{Stage 4: Orienting Matches}
\vspace{-1mm}
\begin{figure}[tb]
    \centering
    \begin{tikzpicture}
        \begin{scope}[every node/.style={minimum size=2em,inner sep=1}]
            \node (R0N)                                                at (0.4,  0) { 0: };
            \node (R00) [draw, circle, fill=uofgpillarbox, text=white] at ($(1.2 * 1,  0)$) { 6 };
            \node (R01) [draw, circle, fill=uofgpillarbox, text=white] at ($(1.2 * 2,  0)$) { 2 };
            \node (R02) [draw, circle, fill=uofglawn]                  at ($(1.2 * 3,  0)$) { 5 };
            \node (R03) [draw, circle, fill=uofglawn]                  at ($(1.2 * 4,  0)$) { 3 };
            \node (R04) [draw, circle, fill=uofgpillarbox, text=white] at ($(1.2 * 5,  0)$) { 4 };
            \node (R05) [draw, circle, fill=uofgpillarbox, text=white] at ($(1.2 * 6,  0)$) { 1 };

@@

The final stage, stage 4, orients skippers within matches, such that they are either port (red) or
starboard (green). This is done to satisfy criteria~2 and~3, 5 and~8. A sketch of the constraint
model is given with a supporting diagram, Figure~\ref{oriented}.  The two dimensional array 
of constrained integer variables $orient[i,j] \in \{0,1\}$ has $orient[i,j] = 0$ if and only if skipper $\sigma[i]$ is on port side in his $j^{th}$ match (starboard otherwise).
For example, taking the schedule in Table~\ref{tab2}, skipper 
$\sigma[{2}]$ meets skippers in the order 3,0,1,4,5,6; skipper 
$\sigma[3]$ meets skippers in the order 2,6,0,5,1,4 and 
$\sigma[{5}]$ meets skippers in the order 4,6,0,1,3,2.  Therefore
$orient[2,0] \neq orient[3,0]$,
$orient[2,4] \neq orient[5,5]$ and
$orient[3,3] \neq orient[5,4]$.
There are summation constraints on each skipper to enforce
criteria~2 and~3 (equal number of port and starboard matches). Criterion~8 is enforced by posting
constraints such that in a match ($\sigma[{i}],\sigma[{j}]$) where $\abs*{i - j} = 1$, the higher 
indexed skipper (lower ranked) is on the port side. A DFA constraint (i.e. \emph{regular}) is posted for criterion~5
(restricting sequences of port and starboard assignments).

@@

\section{Sample Schedules}\label{sec:samples}
\vspace{-2mm}
We present four pairs of schedules. The first pair is for 8 skippers and 6 boats, one published by
ISAF, the other generated by us. Next, we present two new schedules, not appearing in the ISAF
manual, for 9 skippers and 6 boats and 9 skippers with 8 boats, then 10 skippers and 6 boats, again
both published by ISAF and generated by us. Finally, we show what is a relatively easy problem, 8 skippers
and 8 boats.  For each of our schedules (with the exception of the latter ``easy-8'' problem) we used 1 day
CPU time for the first stage (minimizing boat changes) then a further 2 days to minimize imbalance. We were unable to find and prove optimality in this time, therefore some of 
these schedules might be improved. The processor used was an Intel (R) Xeon (R) E5-2660 at 2.20GHz. 
Note, stages 3 and 4 (renaming and orientation) completed in less than a second.

@@

\section{Discussion}
\vspace{-2mm}
It was anticipated that this problem would be easy to model. This naivety was due to the use of the
term ``round-robin'' for these schedules, leading us to believe that we could use some of the models
already in the literature \cite{trick2002,trick2008,henz2004}. This naivety slowly dissolved as
we addressed more and more of the ISAF criteria. A number of models we produced were scrapped due to
misunderstanding about the actual problem, i.e.\ a communication problem between the authors.
Eventually this was resolved by presenting the majority of the criteria as a deterministic finite
automaton. This became our Rosetta Stone, with the surprising benefit that it not only improved the
teams communications, it was also a constraint (the \emph{regular} constraint).

@@

\section{Conclusion}\label{sec:conclusions}
\vspace{-2mm}
\begin{figure}[tb]
    \centering
    \includegraphics[keepaspectratio=true,width=4.1in]{dinghy.jpg}
\caption{Scene from an imaginary dinghy race.}
\label{dinghy}
\end{figure}

>>>>

\section{Introduction}
This work describes a round-robin competition format that arises from sailing, known as match
racing. Competitors, i.e.\ \emph{skippers}, compete against each other in a series of matches taking
place in rounds, known as \emph{flights}.  A match is composed of two skippers, with each skipper in
a boat on their own. Skippers in a match set off together, one on the port side, the other starboard,
and first home wins that match. Criteria published by the International Sailing Federation (ISAF)
dictate what makes a legal match race schedule and what should be optimized. This is a rich set of
criteria, 13 in all, and as far as we know, all published schedules have been produced by hand. This
is a daunting task. A close inspection of these schedules reveals that most are illegal, violating
many of the match race scheduling criteria, many schedules are missing, and some that are legal are
far from the ISAF definition of optimality.

@@

\section{Problem Definition: Round-Robin Pairing Lists}\label{sec:definition}
The guidelines for running match racing events~\cite{isaf} places a number of criteria on how the
competing skippers are scheduled into flights and matches, known as pairing lists. Several of these
criteria are applicable when the number of skippers exceeds the number of available boats, in which
case skippers have to change in and out of boats, and are allowed time in the schedule to check and
fine-tune the boat they change into. Typically, matches set off at 5 minute intervals with each
match taking about 20 minutes. Therefore, if we have 10 skippers and 10 boats we have 45 matches,
9 flights each of 5 matches, and if each flight takes about 50 minutes,  eight or nine
flights a day is a reasonable achievement for most events~\cite{isaf}. Consequently, the number of
boats and skippers involved is typically small.

@@

\section{The Constraint Models}\label{sec:models}
Our approach produces schedules in four stages. The first stage produces a schedule that respects
criteria~1, 4, 11 and~12, and minimizes criteria~10 (boat changes). The second stage constructs a
schedule that respects criteria~1, 4, 11, 12 (again), has criterion~10 as a hard constraint generated
from first stage, and minimizes criterion~6 (balance). This results in a multi-objective schedule that attempts to
minimize boat changes \emph{and} balance. The third stage satisfies criterion~9, and is a simple
translation of the schedule produced in stage 2.  The final stage orients skippers within matches to
address criteria~2, 3, 5 and~8.

@@

\subsection{Stage 1: Minimizing Boat Changes}
\paragraph{Modeling skippers:} The first thing we model, in the first half of
Figure~\ref{model:stage1}, is a skipper. Each skipper $\sigma$ has both a temporal view of their
schedule (``who, if anyone, am I racing in the match at time $t$?''), and a state view (``where am
I racing within flight $f$?'').  The temporal view is via the array $\timeSlot$ defined in
\eqref{var:timeSlot}. If variable $\timeSlot[{i}] = k$ and $k \geq 0$ then this skipper is in a
match with skipper~$k$ at time~$i$.

@@

\subsection{Stage 2: Minimizing Imbalances}
Having produced a schedule that minimizes boat changes, we then minimize imbalance (i.e.\ apply
criterion~6) by extending the model in Figure~\ref{model:stage2}.  Assuming the first stage
produces a schedule with $\beta$ boat changes we post this as a hard constraint \eqref{C11} in stage
2. For each skipper $\sigma$ we introduce a zero/one variable $position[{i,j}]$ \eqref{V9} which is
1 if and only if skipper $\sigma$ is in a match in time $m \cdot j + i$, i.e.\ in the $i^{th}$ match
of the  $j^{th}$ flight \eqref{C12}. Imbalance \eqref{V10} is then computed for each position, as
the absolute amount over or under the expected presence in a given position \eqref{C13}. Finally, we
compute the maximum imbalance over all positions for this skipper (\ref{V11}, \ref{C14}).

@@

\subsection{Stage 3: Renaming Skippers}
Stage 3 then takes as input the schedule produced in stage 2 and renames skippers in the last round
of the last flight to be $(0,1)$, satisfying criterion~9. This is a simple linear process. The
schedule produced for 7 skippers and 6 boats is shown in Table~\ref{tab2}.

\subsection{Stage 4: Orienting Matches}
\begin{figure}[tb]
    \centering
    \begin{tikzpicture}
        \begin{scope}[every node/.style={minimum size=2em,inner sep=1}]
            \node (R0N)                                                at (0.4,  0) { 0: };
            \node (R00) [draw, circle, fill=uofgpillarbox, text=white] at ($(1.2 * 1,  0)$) { 6 };
            \node (R01) [draw, circle, fill=uofgpillarbox, text=white] at ($(1.2 * 2,  0)$) { 2 };
            \node (R02) [draw, circle, fill=uofglawn]                  at ($(1.2 * 3,  0)$) { 5 };
            \node (R03) [draw, circle, fill=uofglawn]                  at ($(1.2 * 4,  0)$) { 3 };
            \node (R04) [draw, circle, fill=uofgpillarbox, text=white] at ($(1.2 * 5,  0)$) { 4 };
            \node (R05) [draw, circle, fill=uofgpillarbox, text=white] at ($(1.2 * 6,  0)$) { 1 };

@@

The final stage, stage 4, orients skippers within matches, such that they are either port (red) or
starboard (green). This is done to satisfy criteria~2,~3, 5 and~8. A sketch of the constraint
model is given with a supporting diagram, Figure~\ref{oriented}.  A two dimensional array 
of constrained integer variables $orient[i,j] \in \{0,1\}$ has $orient[i,j] = 0$ if and only if skipper $\sigma[i]$ is on port side in his $j^{th}$ match (starboard otherwise).
For example, taking the schedule in Table~\ref{tab2}, skipper 
$\sigma[{2}]$ meets skippers in the order 3,0,1,4,5,6; skipper 
$\sigma[3]$ meets skippers in the order 2,6,0,5,1,4 and 
$\sigma[{5}]$ meets skippers in the order 4,6,0,1,3,2.  Therefore
$orient[2,0] = 1 ~and~ orient[3,0] = 0$ (criterion 8),
$orient[2,4] \neq orient[5,5]$ and
$orient[3,3] \neq orient[5,4]$.
Summation constraints on each skipper enforce
criteria~2 and~3 (equal number of port and starboard matches). Criterion~8 is enforced by posting
constraints such that in a match ($\sigma[{i}],\sigma[{j}]$) where $\abs*{i - j} = 1$, the higher 
indexed skipper (lower ranked) is on the port side. A DFA constraint (i.e. \emph{regular}) is posted for criterion~5
(restricting sequences of port and starboard assignments).

@@

\section{Sample Schedules}\label{sec:samples}
We present four pairs of schedules. The first pair is for 8 skippers and 6 boats, one published by
ISAF, the other generated by us. Next, we present two new schedules, not appearing in the ISAF
manual, for 9 skippers and 6 boats and 9 skippers with 8 boats, then 10 skippers and 6 boats, again
both published by ISAF and generated by us. Finally, we show what is a relatively easy problem, 8 skippers
and 8 boats.  For each of our schedules (with the exception of the latter ``easy-8'' problem) we used 1 day
CPU time for the first stage (minimizing boat changes) then a further 2 days to minimize imbalance. We were unable to find and prove optimality in this time, therefore some of 
these schedules might be improved. The processor used was an Intel (R) Xeon (R) E5-2660 at 2.20GHz. 
Note, stages 3 and 4 (renaming and orientation) completed in less than a second.

@@

\section{Discussion}
It was anticipated that this problem would be easy to model. This naivety was due to the use of the
term ``round-robin'' for these schedules, leading us to believe that we could use some of the models
already in the literature \cite{trick2002,trick2008,henz2004}. This naivety slowly dissolved as
we addressed more and more of the ISAF criteria. A number of models we produced were scrapped due to
misunderstanding about the actual problem, i.e.\ a communication problem between the authors.
Eventually this was resolved by presenting the majority of the criteria as a deterministic finite
automaton. This became our Rosetta Stone, with the surprising benefit that it not only improved the
teams communications, it was also a constraint (the \emph{regular} constraint).

@@

\section{Conclusion}\label{sec:conclusions}
\begin{figure}[tb]
    \centering
    \includegraphics[keepaspectratio=true,width=4.1in]{dinghy.jpg}
\caption{Scene from an imaginary dinghy race.}
\label{dinghy}
\end{figure}
